---
title: "Notes_10_29"
output: html_document
date: "2024-10-29"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

Comparing methods:
  A randomization distribution shows us EXACTLY how the sample statistics
vary under Null, All other methods are only approx.

```{r}

```

When we use Normal Distribution to calculate the p-value [N(0,1), or N(null, SE)] we are assuming that EXACT randomization distribution can be well approximated with a normal distribution.

If our sample size(n) is too small, then the approx is NOT good, and the p-value using a Normal approx. is WRONG.

On a day with 21 students present 19 of them had iPhones:

  Proportion of randomization distr, stats > 19/21 is 0.079.
  
  Area under the Normal curve, 0.051.
  

What's realy happening with t-distrution:
  Our goal is to use some example to really understand CT and approximation with a distribution.
  can't use with small sample size that is skewed. Because look at bootstrap distribution and check.
  Use t because we don't have standard deviation, because the standard error is not as accurate for approximationg p-value.
  If you do have standard deviation then use normal distribution
  Remember using that using s instead of sigma introduces extra problems.
  
  