---
title: 'Lab 1-2: Using R Markdown to explore data'
output: html_notebook
---

### Objectives:

For this lab you should...

-   Read datasets into R
-   Familiarize yourself with commonly used functions and model-centric syntax in R
-   Provide summary statistics of a dataset
-   visualize the dataset using histograms and boxplots

# Part 1: Getting ready to use data

The following packages will need to be loaded into R before you begin the lab. To run a chunk of R code, click the green arrow symbol in the upper right of the gray chunk. If you would like to run line by line, hit Ctrl-Enter with your cursor next to the desired line.

```{r, echo = F, message = F}
##  NOTE - in all future documents/assignments this code will be included for you and you are expected to run it without prompting.


# Clear Workspace
rm(list = ls()) 
knitr::opts_chunk$set(echo=FALSE)

# load packages we typically use for this class.
library(mosaic, warn.conflicts = FALSE) 
library(ggformula, warn.conflicts = FALSE)


```

### The Happiness Data

The happiness data set is from the 2021 World Happiness Report <https://Worldhappiness.report/ed/2021>. Each observation in this dataset represents a country in the world. There are 14 variables observed about citizens opinions on their countries, characteristics of their countries, and Covid-19 related summaries. A multitude of interesting questions can be posed (and answered!) using these data.

**TASK 1.0** Download the dataset WorldHappiness.csv from the Lab 1.1 assignment page. Once you've downloaded it, using the following code to read it into R:

```{r,echo = F}
# The code below reads in the data as a dataframe and stores it in R using the name 'Hap'

Hap <- read.csv(file.choose()) #select your file in its saved location when the prompt opens

```

**TASK 1.1** To ensure that you have successfully loaded the dataset, you can check using the function head(DataSetName), which displays the first six rows of the data, and glimpse(DataSetName), which provides a summary of the data frame.

```{r,echo = F}
# modify the code below

head(Hap) # displays the first 6 rows of the dataset, including headers 
glimpse(Hap) # provides a summary of the data frame

```

**TASK 1.2** Use your results from Task 1.1 to determine which columns are categorical data, and which are numeric:

```         
**Response**

Categorical: Name, Region, FemaleGovBoss, Island

Numeric: HappinessScore, Population2019, CovidDeaths2020, MedianAge, InstitutionalTrust, ExcessDeaths, LogGDP
SocialSupport, HealthyLifeExpectancy, Freedom, PerceptionsOfCorruption
```

### Model-centric approach to R

For the purposes of our class, it's useful to learn a model-centric approach to R. The psuedo-code below is going to be our foundation for the rest of the class:

`function( Y ~ X, data = DataSetName )`

Here's a short description of each part in the pseudo-code above:

-   `function` is an R function that dictates something you want to do with your data e.g.
    -   for example, `mean` calculates the mean
-   `Y` is the outcome of interest (response variable)\
-   `X` is some explanatory variable or you can use "1" as a placeholder if there is no explanatory variable\
-   `DataSetName` is the name of a data set loaded into the R environment

A summary of the most commonly used functions for this class are provided at the bottom of this document. In the work below you will be introduced to the functions and their uses one at a time in the context of answering research questions.

## Research Question A: Are different regions of the world happier?

To use the data to answer the research question, we could find data summaries such as average happiness, for each of the regions. In this situation, HappinessScore is our response variable and Region is our explanatory variable.

**TASK 2.1** Modify and run the code below to calculate the mean HappinessScore for each region.

```{r,echo = F}

mean(HappinessScore ~ Region, data = Hap)

```

**TASK 2.2** Use the function 'favstats()', and model-centric syntax, to determine which region has the country with the highest happiness score.

```         
**Response:** 
Western Europe
```

```{r, echo = F}

  favstats(HappinessScore ~ Region, data=Hap)

```

**TASK 2.3** Which summary statistics are calculated by the favstats() function?

```         
**Response:**
Minimum score, Maximum Score, Median, Mean, 1st quarter percentile, 3rd quarter percentile, standard deviation, 
number of entries for each category of the explanatory variable, how many missing response variable values exist for each category of the explanatory variable.
```

Now it's time to visualize the distribution of Happiness Score for the different reasons.

**TASK 2.4** Modify the code below to create a histogram of Happiness Score, without accounting for other variables. What is the shape of the distribution: left-skewed, right-skewed, or roughly symmetric?

```         
**Response:**
Roughly Symmetric
```

```{r,echo = F}

gf_histogram( ~HappinessScore, data = Hap)

```

**TASK 2.5** Generate side-by-side boxplots of the HappinessScore, separated by Region. Write a sentence about what you observe in the figure.

```         
**Response:**
```

More developed regions tend to have higher happiness scores than less developed regions.

```{r,echo = F}
#gf_boxplot(HappinessScore ~ Region, data = Hap)

# if the axes labels aren't readable, try modifying and running the code below!
gf_boxplot(HappinessScore ~ Region, data = Hap) + coord_flip()

```

**TASK 2.6** Based on everything you've learned, do you think that happiness scores vary by region?

```         
**Response:**
Yes
```

#### Research Question B: Are countries run by female leaders happier?

**TASK 3.1** Write and run code to determine whether the mean Happiness Score is higher or lower for countries with female leaders. Hint: the variable 'FemaleGovBoss' indicates whether a country is run by a female leader.

```         
**Response:**
```

Countries with a female leader have a higher happinessScore on Average.

```{r,echo = F}

  mean(HappinessScore ~ FemaleGovBoss, data=Hap)

```

**TASK 3.2** How many countries are governed by female leaders? You can tabulate observations of a categorical variable using the 'tally' command. Modify the code below to answer the question.

```         
**Response:**
22
```

```{r,echo = F}
tally( ~ FemaleGovBoss, data = Hap)
```

**TASK 3.3** Create an appropriate visual summary to investigate research question B.

```{r,echo = F}

  gf_boxplot( HappinessScore ~ FemaleGovBoss, data=Hap) + coord_flip()

```

**TASK 3.4** Based on what you learned in tasks 3.1 and 3.3, what do you think the answer to research question B would be?

```         
**Response:**
The answer is yes
```

## Research Question C: Are Island nations more likely to have female leaders?

**TASK 4.1** Use R tally up how many countries are islands with female leaders, islands are male leaders, not islands with female leaders, and not islands with male leaders. Modify the code below to extract the numbers.

```{r, echo = F}
tally(FemaleGovBoss ~ Island, data = Hap)
```

For a research question like this one that involves two binary variables, it is useful to calculate and compare two proportions:

**TASK 4.2** What proportion of Island Nations have female leaders? You should calculate (\# island nations with female leaders)/(\# island nations)

```         
**Response:**
0.15
```

```{r}
# you can use this R chunk as a calculator if needed
(3)/(20)

```

**TASK 4.3** What proportion of non-Island Nations have female leaders? You should calculate (\# non-island nations with female leaders)/(\# non-island nations)

```         
**Response:**
0.1484375
```

```{r}
# you can use this R chunk as a calculator if needed
19/(109+19)
```

To visualize the data as it relates to this question, you should make side-by-side or stacked barcharts.

**TASK 4.4** Modify the code below to create a side by side barchart. Notice the syntax is a little different; you should try different options for X and Y until you have Islands on the bottom and different colors for FemaleGovBoss.

```{r, echo = F}
gf_bar(~Island, fill =  ~ FemaleGovBoss, data = Hap, position = position_dodge())
```

**TASK 4.5** Based on your answers to Tasks 4.1--4.4, what do you think the answer to research question C would be?

```         
**Response:**
They are a little bit more likely to have a female leader, but the chances are pretty much equal.
```

## Research Question D: Which of the quantitative variables are related to Happiness Score?

To answer this question, we consider plots that visualize the relationship between two quantitative variables. A scatterplot is simply an (x,y) plot in the Cartesian plane. Using our Y \~ X format, the Y variable will be along the Y axis.

**TASK 5.1** Modify the code below to generate a scatterplot of the HappinessScore versus the Institutional Trust.

```{r,echo = F}
gf_point(HappinessScore ~ InstitutionalTrust, data = Hap)

```

**TASK 5.2** Write code and produce scatterplots of the HappinessScore versus: SocialSupport, HealthyLifeExpectancy, and Freedom.

```{r,echo = F}

gf_point(HappinessScore ~ SocialSupport, data = Hap)
gf_point(HappinessScore ~ HealthyLifeExpectancy, data = Hap)
gf_point(HappinessScore ~ Freedom, data = Hap)

```

**TASK 5.3** Comment on which of the four variables that have been plotted suggest a strong linear relationship with the Happiness Score.

```         
**Response** 
Healthy Life Expectancy has the stongest linear relationship of the the four variables with Social Support being the next best linear relationship.
```

**Response**

**TASK 5.4** Calculate the correlation for each of the four variable pairs above.

```         
**Response** 

HappinessScore and InstitutionalTrust:0.0709448

HappinessScore and SocialSupport: 0.7576146

HappinessScore and HealthyLifeExpectancy: 0.7667235

HappinessScore and Freedom: 0.6038699
```

```{r, echo = F}
cor(HappinessScore ~ InstitutionalTrust, data=Hap)
cor(HappinessScore ~ SocialSupport, data=Hap)
cor(HappinessScore ~ HealthyLifeExpectancy, data=Hap)
cor(HappinessScore ~ Freedom, data=Hap)

```

## Research Question E: Which variables are related to Covid Deaths 2020?

A strength of R is its ability to quickly and easily produce sophisticated data visualizations. You will use curiosity and tools in R to create visualizations that explore this research question.

Use the code below to remind yourself which variables are available in this dataset. Choose a quantitative variable that you think will be related to CovidDeaths2020.

**TASK 6.1** Make a scatterplot to examine the relationship between Covid Deaths in 2020 and the variable you chose.

```{r,echo = F}
gf_point(CovidDeaths2020 ~ HealthyLifeExpectancy, data=Hap)

```

**TASK 6.2** Now, color the points in your plot based on the values of a third variable such as FemaleGovBoss, Island, or Region. You could also choose a quantitative variable, if you desired.

```{r,echo = F}

gf_point(CovidDeaths2020 ~ HealthyLifeExpectancy, col = ~ MedianAge, data = Hap)


```

**TASK 6.3** Now we can add one more layer, based on a categorical variable. Choose one of the binary variables FemaleGovBoss or Island that you haven't already included and modify the code below:

```{r, echo = F}
gf_point(CovidDeaths2020 ~ HealthyLifeExpectancy | Island, col = ~MedianAge, data = Hap)
```

**TASK 6.4** Make two interesting comments about what you learned about the variables you chose and covid deaths in 2020.

```         
**Response:**
```

# Epilogue

Below are some summaries you can reference at any time.

#### Numerical summaries

The primary numerical summaries used in this course include the mean, the five-number summary, and the standard deviation. The functions below calculate these summaries for a single variable, or for that variable based on the value of a second categorical variable.

-   mean(\~ Y, data = DataSetName) \# mean of Y without considering other variables
-   mean(Y \~ X, data = DataSetName) \# mean of Y for different values of X
-   sd(\~ Y, data = DataSetName) \# standard deviation of Y
-   sd(Y \~ X, data = DataSetName) \# standard deviation of Y for different values of X
-   favstats(\~ Y, data = DataSetName) \# summary statistics of Y
-   favstats(Y \~ X, data = DataSetName) \# summary statistics of Y for different values of X

#### Visual summaries

The primary graphical representations used in this course will be a barcharts, boxplots, histograms and scatterplots. Each plot helps to visualize different aspects of a dataset.

The functions that will be used to generate these plots are:

-   gf_boxplot(Y \~ X, data = DataSetName) \# boxplot of Y for each value of X, if specified
-   gf_histogram(\~Y, data = DataSetName) \# histogram of Y
-   gf_point(Y \~ X, data = DataSetName) \# scatterplot of Y versus X
-   gf_bar( \~Y, data = DataSetName) \# Bar chart for a single variable
-   gf_bar( \~ X, fill = \~ Y, data = DataSetName, position = position_dodge()) \# side-by-side bar chart
-   gf_bar( \~ X, fill = \~ Y, data = DataSetName) \# stacked bar chart
